PyTracer: Automatically Profiling Numerical Instabilities in Python

### 2022-10-17 IEEEtran/paper.tex
Remove \add of first review

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
    \begin{abstract}
        Numerical stability is a crucial requirement of reliable scientific
        computing. However, despite the pervasiveness of Python in data science,
        analyzing large Python programs remains challenging due to the lack of
        scalable numerical analysis tools available for this language. To fill
        this gap, we developed \pytracer, a profiler to quantify numerical
        instability in Python applications. \pytracer transparently instruments
        Python code to produce numerical traces and visualize them interactively
        in a Plotly dashboard. We designed \pytracer to be agnostic to numerical
        noise model, allowing for \add{numerical profiling} through Monte-Carlo
        Arithmetic, random rounding, random data perturbation, or structured
        noise for a particular application. We illustrate \pytracer's
        capabilities by testing the numerical stability of key functions in both
        SciPy and Scikit-learn, two dominant Python libraries for mathematical
        modeling. Through these evaluations, we demonstrate \pytracer as a
        scalable, automated, and generic framework for numerical profiling in
        Python.
    \end{abstract}

@@

\pytracer estimates the variability of floating-point variables by
combining program traces obtained with different numerical perturbations,
which requires (\add{a}\remove{1}) generating numerically-perturbed
executions, (\add{b}\remove{2}) tracing floating-point computations along with
program executions, and (\add{c}\remove{3}) visualizing stability
evaluations. \add{PyTracer addresses this work (a) by relying on
    Fuzzy~\cite{kiar2020comparing}, an existing collection of scientific
    computing libraries instrumented with Monte-Carlo arithmetic through the
    Verificarlo tool~\cite{verificarlo}, (b) by dynamically instrumenting
    Python functions, modules, and classes through meta-programming, and (c) by
    visualizing stability measures throughout program executions with an
    interactive Plotly dashboard. \mbox{\pytracer} is available at
    \mbox{\href{https://github.com/big-data-lab-team/pytracer}{github.com/big-data-lab-team/pytracer}}
    under MIT License.} \remove{\mbox{\pytracer} addresses these requirements
    with (1) a Fuzzy Python interpreter instrumented with Monte-Carlo
    arithmetic, (2) a dynamic instrumentation of Python functions, modules, and
    classes through meta-programming, (3) an interactive Plotly dashboard to
    visualize stability measures throughout program executions. }

This manuscript presents the design of \pytracer and \add{highlights} its potential
in various use cases. Through stability evaluations of the SciPy and
scikit-learn libraries, we demonstrate \pytracer as a practical solution to
review the numerical stability of extensive code bases.

@@

\add{
    To detect numerical instabilities, \pytracer works in three main steps:
    \begin{enumerate}
        \item It instruments Python programs and their module dependencies
              automatically, to save the input and output values of functions in a trace
              file.
        \item It executes the instrumented program several times using Monte-Carlo
              Arithmetic to inject perturbations in function outputs.
        \item It aggregates all trace files to compute statistics about the traced
              values, and visualize them in a GUI.
    \end{enumerate}
    \pytracer adopts the workflow shown in Figure~\ref{fig:workflow}.
    In Step 1, \pytracer uses monkey patching to replace the functions called
    by the application with a wrapper that saves their arguments and returned
    values in a trace file. In Step 2, it executes the application multiple
    times with Monte-Carlo Arithmetic as implemented in the
    Fuzzy~\cite{kiar2020comparing} environment. It then
    aggregates the resulting trace files to compute summary statistics on
    traced variables. In Step 3, a Plotly dashboard provides interactive
    visualizations highlighting unstable code sections.
}

\add{
    The remainder of this section assumes some knowledge of Python concepts
    such as classes, functions, modules, decorators, callables, and iterators.
    We refer to~\cite{beazley2013python} for a detailed presentation of these
    concepts.
}

@@

\pytracer dynamically instruments Python modules without requiring source code
modifications in the application. To do so, it creates a new, instrumented
version of each module that preserves its original attributes so that the
application can transparently use it.
% \gkmod{}{Note: this form of monkey patching is basically global and
% transparent decorating, right?} \Yohan{Yes}
By default, \pytracer instruments all the attributes in a given module except
the special ones of the form \texttt{\_\_*\_\_}, which is common for reserved
Python methods. However, some attributes are not writable. In this case,
\pytracer \add{preserves} the original attributes and warns
the user. The user can also restrict the list of traced attributes through an
inclusion/exclusion mechanism. The main advantage of this instrumentation
approach is the lack of required modification in the application source code, in
contrast with \remove{decorator-based approaches} \add{a decorator-based
    approach~\mbox{\cite{hunt2019decorators}}}. In addition, this technique is
scalable and applicable in read-only environments such as Singularity containers
or HPC clusters.

@@

\remove{Listings~\mbox{\ref{fig:wrapper_creation}} and~\mbox{\ref{fig:generic_wrapper}} show
    \mbox{\pytracer}'s function instrumentation mechanism.}
For each function to be instrumented, \pytracer's \texttt{loader} class creates a
wrapper function, dynamically compiles it with the \texttt{compile} builtin
function, and substitutes it for the original module function.  Although simple,
this technique does not apply to callable class instances, i.e., class instances
\add{that can be called directly with the () operator, because they} are not of type
\texttt{FunctionType}. Indeed, applying the wrapper technique would 1) modify
the type of the class instance to \texttt{FunctionType}, which could cause
syntax and semantic errors, and 2) mask class attributes, leading to
\texttt{AttributeError} exceptions. \pytracer overrides the
\texttt{\_\_call\_\_} attribute with the wrapper function when possible to
overcome this issue. When the \texttt{\_\_call\_\_} attribute is not writable,
Pytracer does not instrument the class and returns a warning. 
\add{
    This limitation is expected to have a minor impact in practice since non-writable
    classes are usually called from pure Python classes that can be
    instrumented. Therefore, \pytracer reports information about the stability of
    non-writable classes through the instrumentation of the pure Python
    classes that use them.
}

@@

\add{
    We do not expect this limitation to have a critical impact on the
    applicability of \pytracer. 
    In general, iterators involve simple computations that are unlikely to be
    unstable. In particular, we encountered only a few iterators in our
    experiments and none of them involved floating-point values. Should
    instabilities occur in iterators, \pytracer would detect them through the
    inputs of the functions consuming these iterators. 
}

@@

Three  main  approaches  exist  to  detect  numerical  instabilities:
stochastic  arithmetic,  uncertainty  or  sensitivity analysis~\cite{helton2006survey}, and random
seed analysis~\cite{hellekalek1998good}. \add{PyTracer is agnostic to the noise model used and can therefore
    be used for all these methods}.  For our use cases, we focus on stochastic
arithmetic since it does not make any assumption about the noise model, in
contrast to sensitivity analysis\add{.}

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing for simulating
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
\add{In our experiments, we used MCA in Random Rounding (RR) mode, which
    simulates roundoff errors by applying function $inexact$ only to the
    output returned by a function. The Precision Bounding (PB) mode that applies
    the $inexact$ function on inputs can
    identify other types of instabilities, in particular catastrophic
    cancellations, however, it is more difficult to apply to large code bases
    as it often leads to execution errors that require human intervention.}

Stochastic arithmetic quantifies numerical error using the number of significant
digits $s$ estimated among the sampled values. A common formula to determine
this number from MCA samples is presented in~\cite{parker1997monte}:
\begin{equation}
    s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a
variable sampled with MCA. \add{This can be seen as the ratio of signal over noise.}
Sohier et al.~\cite{sohier2021confidence} recently provided a generalization of
this formula to include confidence intervals.

@@

\add{Monte Carlo Arithmetic samples the floating-point variability of a given
    program, and hence the more MCA repetitions we have, the more accurate the
    distribution will be. In theory, an infinite number of MCA repetitions
    should give the exact distribution. While in practice low sample sizes (3 to 30)
    can give good estimates, the issue is to capture ``rare events" that could make the
    computation strongly diverge from the mean. Determining the number of
    MCA samples to use for a given instrumented code is still an open research question. 
%%     Recent theoretical works}~\cite{sohier2021confidence}
%% \add{on the estimation of significant digits allow computing an associated
%%   confidence interval depending on the sample size and the underlying distribution.
}

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information is loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. It assumes that the application is
single-threaded and that the control flow does not depend on a random state.
\add{Therefore, \mbox{\pytracer} cannot merge traces resulting from different
    control flow paths and will stop the parsing when encountering such a situation.}

@@

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA through Fuzzy in the Python interpreter \add{as well as} BLAS, LAPACK, NumPy,
SciPy, and scikit-learn. \add{These libraries are written in large parts in a
    compiled language other than Python (such as C or Cython) and their execution is not 
    perturbed by the sole use of Fuzzy Python. We chose to activate MCA in these libraries 
    to provide an overall picture of numerical uncertainty in the tested examples. 
    Depending on the use case of interest, one might choose to use MCA only in 
    a subset of these libraries or even just in the Python interpreter.} \add{We
    used MCA in Random Rounding (RR) mode, with} the default virtual
precision in Verificarlo, namely 24 bits for single-precision floating-point
values and 53 bits for double-precision values, which simulates machine error.
We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.

\begin{table*}[]
    \centering
    \normalsize
    \begin{subfigure}[t]{\linewidth}
        \centering
        \begin{tabular}{|lll|c|c|c|}
            \multicolumn{3}{|c}{ \multirow{2}{*}{Application} }                 & \multicolumn{3}{|c|}{Random Rounding (RR)}                                                                                                                         \\
            \cline{4-6}
                                                                                &                                                                           &                                                          & trace  & parse    & sigbits \\
            \multicolumn{1}{|c|}{ \multirow{20}{2em}{ \rotatebox{90}{SciPy} } } & 
            \multicolumn{1}{c|}{ \multirow{3}{2em}{ \rotatebox{90}{FFT} }}      & \discreteCosineRF                                                         & \valid                                                   & \valid & \s{52}             \\ 
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \fftOneDRf                                               & \valid & \valid   & 41      \\
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \fftTwoDRf                                               & \valid & \valid   & \s{52}  \\
            \cline{2-6}
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{  \multirow{5}{*}{ \rotatebox{90}{Interpolation} } }
                                                                                & \interOneDRf                                                              & \valid                                                   & \valid & 51                 \\
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \multiRf                                                 & \valid & \valid   & \s{51}  \\
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \bsplineRf                                               & \valid & \valid   & 10      \\
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \splineOneDRf                                            & \valid & \cross   & \s{52}  \\
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \splineTwoDRf                                            & \valid & \valid   & 10      \\
            \cline{2-6}
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{ \multirow{12}{2em}{ \rotatebox{90}{Optimization} } }

@@

        % \tristan{Replace FFT with SciPy FFT, Interpolation with SciPy
        % Interpolation and Optimization with SciPy Optimization, or add a SciPy
        % header to these rows.} \tristan{Nitpick: you could add a color scheme
        % to sigbits (for instance shades of blue), to make the column more
        % easily readable.} \\

@@

        \pytracer trace parsing and aggregation, \emph{result}: \add{result} precision
        in number of significant bits. \valid: successful run, \remove{\warn: error
            from invalid noise injection,} \cross: numerical error \add{(tracing)} or 
        \add{different control flow paths (parsing).} \textit{nan} stands for (\texttt{Not-A-Number}). Example names
        are linked to their source code.}

@@

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on
SciPy and scikit-learn examples. Columns \textit{trace} and \textit{parse}
report on the tracing and aggregation steps with two possible outcomes:
\add{green \valid\xspace for success and red \cross\xspace for error. Tracing errors
    occur in the application as a result of 
    numerical perturbations, whereas parsing errors occur when traces result from different 
    control flow paths.}
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random
Rounding mode was entirely successful for 32/40 examples\remove{ and showed an average
    precision loss of 8 bits}. 
\remove{The instrumentation with Full MCA is
    more invasive and reduced the number of successful exe-
    cutions to 12/40. Among the 28/40 failed executions, one
    failed due to an actual numerical error (Bayesian Ridge
    Regression example) and the other ones failed due to
    perturbations that should not have been introduced.}

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the \add{result} precision for the main outputs of SciPy and scikit-learn examples. We
measured the \add{result} precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produce\add{d multiple} outputs, the average value among them. \add{Significant bits were computed on the final program output
    and could therefore be computed even when parsing failed.} \add{We use the term ``result precision"
    to denote the statistical variability measured in the number of significant bits, in 
    contrast with the other meanings of the word precision used in the expressions
    ``virtual precision" or ``single/double precision'' in this paper.} The numerical
quality across all examples varied from 0 to 52, with an average of 44
significant bits. We note that only one example failed due to a
numerical error (\texttt{face\_recognition}
\remove{for RR and \texttt{Bayesian Ridge Regression} for full MCA}).

@@

\add{PyTracer supports native floating-point formats as well as compound
    data types, such as complex floating-point values, N-dimensional arrays,
    and arbitrary Python classes.} Figures~\ref{fig:fft1D_inputs}
and~\ref{fig:fft1D_outputs} \add{show the visualization of the
    \texttt{fft\_1D} input and output values. Results are complex values
    encoded in double precision. PyTracer allows visualizing the
    \add{result} precision of real (blue points) and imaginary (orange
    points) parts.} \add{The} top row in Figures~\ref{fig:fft1D_inputs}
and~\ref{fig:fft1D_outputs} show the input mean value while the bottom
row shows the standard deviation. \add{Even though Random Rounding
    (RR) was used, function inputs were perturbed since they result from
    multiple intermediate computations perturbed with RR.} The points with
low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs
near 0 when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in
    \mathbb{Z}$. Indeed, since MCA introduces noise, the result is not
exactly 0. We can see in Figure~\ref{fig:fft1D_inputs} that the maximal
noise introduced by MCA on the inputs is in the order of $10^{-14}$,
two orders of magnitude higher than the $ulp \simeq 10^-16$ for double precision. This slight
noise introduced can be interpreted as white noise on the input signal. We can
see on the bottom row of Figure~\ref{fig:fft1D_outputs} that the frequency noise
is of the same order of magnitude as the input noise, which is expected. The
two peaks at x=38 and x=562 correspond to the actual frequencies of the input
signal.

Figure~\ref{fig:spline2d_rr} \add{shows the results of the spline evaluation with
    \texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
    interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
        [-1,1]\times[-1,1]$. The result is a 2D array that is natively supported by
    PyTracer. We observe a significant loss of \add{result} precision in some regions, which follows
    an interesting pattern that could be further investigated.
    This example shows that \pytracer workds on 2D inputs and the importance of
    visualizing complex data and not only reporting the values.}

@@

\add{We illustrate on two examples how PyTracer is effective for analyzing
    numerical stability in complex codes at a fine-grained level: Face recognition and Bayesian Ridge Regression.}
The face recognition example uses eigenfaces (eigenvectors) and Support Vector
Machines (SVMs) to classify faces from the Labeled Faces in the Wild dataset.
This example uses a Principal Component Analysis (PCA) to extract the eigenfaces
and then trains an SVM classifier to predict the label associated with a face.
\add{When executing in PyTracer, this example crashed because the Singular Value Decomposition (SVD) did not converge.}
This example is the only one to raise an error due to the
non-convergence of the SVD performed in the PCA.
The PCA uses the Randomized SVD~\cite{halko2011finding} method
that first computes a low-rank approximation of the input matrix using a
stochastic algorithm and then uses this approximation as input for the SVD. 
\add{When walking through the code with PyTracer, we observed that}
the \texttt{randomized\_svd} scikit-learn function uses the Divide \& Conquer SVD
method \texttt{gesdd}. The LAPACK library has two main methods for computing the SVD:
\texttt{gesdd} and \texttt{gesvd}. The former uses a Divide \& Conquer approach,
while the latter uses a QR decomposition. While both methods are expected to
have the same accuracy~\cite{nakatsukasa2013stable}, the Divide \& Conquer
does not converge with MCA.

>>>>

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
    \begin{abstract}
        Numerical stability is a crucial requirement of reliable scientific
        computing. However, despite the pervasiveness of Python in data science,
        analyzing large Python programs remains challenging due to the lack of
        scalable numerical analysis tools available for this language. To fill
        this gap, we developed \pytracer, a profiler to quantify numerical
        instability in Python applications. \pytracer transparently instruments
        Python code to produce numerical traces and visualize them interactively
        in a Plotly dashboard. We designed \pytracer to be agnostic to numerical
        noise model, allowing for numerical profiling through Monte-Carlo
        Arithmetic, random rounding, random data perturbation, or structured
        noise for a particular application. We illustrate \pytracer's
        capabilities by testing the numerical stability of key functions in both
        SciPy and Scikit-learn, two dominant Python libraries for mathematical
        modeling. Through these evaluations, we demonstrate \pytracer as a
        scalable, automated, and generic framework for numerical profiling in
        Python.
    \end{abstract}

@@

\pytracer estimates the variability of floating-point variables by
combining program traces obtained with different numerical perturbations,
which requires (a) generating numerically-perturbed
executions, (b) tracing floating-point computations along with
program executions, and (c) visualizing stability
evaluations. PyTracer addresses this work (a) by relying on
Fuzzy~\cite{kiar2020comparing}, an existing collection of scientific
computing libraries instrumented with Monte-Carlo arithmetic through the
Verificarlo tool~\cite{verificarlo}, (b) by dynamically instrumenting
Python functions, modules, and classes through meta-programming, and (c) by
visualizing stability measures throughout program executions with an
interactive Plotly dashboard. \mbox{\pytracer} is available at
\mbox{\href{https://github.com/big-data-lab-team/pytracer}{github.com/big-data-lab-team/pytracer}}
under MIT License.

This manuscript presents the design of \pytracer and highlights its potential
in various use cases. Through stability evaluations of the SciPy and
scikit-learn libraries, we demonstrate \pytracer as a practical solution to
review the numerical stability of extensive code bases.

@@

To detect numerical instabilities, \pytracer works in three main steps:
\begin{enumerate}
    \item It instruments Python programs and their module dependencies
          automatically, to save the input and output values of functions in a trace
          file.
    \item It executes the instrumented program several times using Monte-Carlo
          Arithmetic to inject perturbations in function outputs.
    \item It aggregates all trace files to compute statistics about the traced
          values, and visualize them in a GUI.
\end{enumerate}
\pytracer adopts the workflow shown in Figure~\ref{fig:workflow}.
In Step 1, \pytracer uses monkey patching to replace the functions called
by the application with a wrapper that saves their arguments and returned
values in a trace file. In Step 2, it executes the application multiple
times with Monte-Carlo Arithmetic as implemented in the
Fuzzy~\cite{kiar2020comparing} environment. It then
aggregates the resulting trace files to compute summary statistics on
traced variables. In Step 3, a Plotly dashboard provides interactive
visualizations highlighting unstable code sections.

The remainder of this section assumes some knowledge of Python concepts
such as classes, functions, modules, decorators, callables, and iterators.
We refer to~\cite{beazley2013python} for a detailed presentation of these
concepts.

@@

\pytracer dynamically instruments Python modules without requiring source code
modifications in the application. To do so, it creates a new, instrumented
version of each module that preserves its original attributes so that the
application can transparently use it.
% \gkmod{}{Note: this form of monkey patching is basically global and
% transparent decorating, right?} \Yohan{Yes}
By default, \pytracer instruments all the attributes in a given module except
the special ones of the form \texttt{\_\_*\_\_}, which is common for reserved
Python methods. However, some attributes are not writable. In this case,
\pytracer preserves the original attributes and warns
the user. The user can also restrict the list of traced attributes through an
inclusion/exclusion mechanism. The main advantage of this instrumentation
approach is the lack of required modification in the application source code, in
contrast with  a decorator-based approach~\mbox{\cite{hunt2019decorators}}. 
In addition, this technique is
scalable and applicable in read-only environments such as Singularity containers
or HPC clusters.

@@

\remove{Listings~\mbox{\ref{fig:wrapper_creation}} and~\mbox{\ref{fig:generic_wrapper}} show
    \mbox{\pytracer}'s function instrumentation mechanism.}
For each function to be instrumented, \pytracer's \texttt{loader} class creates a
wrapper function, dynamically compiles it with the \texttt{compile} builtin
function, and substitutes it for the original module function.  Although simple,
this technique does not apply to callable class instances, i.e., class instances
that can be called directly with the () operator, because they are not of type
\texttt{FunctionType}. Indeed, applying the wrapper technique would 1) modify
the type of the class instance to \texttt{FunctionType}, which could cause
syntax and semantic errors, and 2) mask class attributes, leading to
\texttt{AttributeError} exceptions. \pytracer overrides the
\texttt{\_\_call\_\_} attribute with the wrapper function when possible to
overcome this issue. When the \texttt{\_\_call\_\_} attribute is not writable,
Pytracer does not instrument the class and returns a warning. 
This limitation is expected to have a minor impact in practice since non-writable
classes are usually called from pure Python classes that can be
instrumented. Therefore, \pytracer reports information about the stability of
non-writable classes through the instrumentation of the pure Python
classes that use them.

@@

We do not expect this limitation to have a critical impact on the
applicability of \pytracer. 
In general, iterators involve simple computations that are unlikely to be
unstable. In particular, we encountered only a few iterators in our
experiments and none of them involved floating-point values. Should
instabilities occur in iterators, \pytracer would detect them through the
inputs of the functions consuming these iterators.

@@

Three  main  approaches  exist  to  detect  numerical  instabilities:
stochastic  arithmetic,  uncertainty  or  sensitivity analysis~\cite{helton2006survey}, and random
seed analysis~\cite{hellekalek1998good}. PyTracer is agnostic to the noise model used and can therefore
be used for all these methods.  For our use cases, we focus on stochastic
arithmetic since it does not make any assumption about the noise model, in
contrast to sensitivity analysis.

Stochastic arithmetic leverages randomness to estimate numerical instabilities
coming from the use of floating-point representations. The main idea is to treat
round-off errors as random variables and characterize them statistically. Monte 
Carlo Arithmetic (MCA)~\cite{parker1997monte} uses this principle by introducing
two improvements: (i) a virtual precision parameter, allowing for simulating
reduced working precision, and (ii) different perturbation modes to introduce
perturbations on function inputs or output. MCA replaces each floating-point
number $x$ with a stochastic counterpart computed as:
\[
    inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $e_x$ is the magnitude of $x$, $t$ the virtual
precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
In our experiments, we used MCA in Random Rounding (RR) mode, which
simulates roundoff errors by applying function $inexact$ only to the
output returned by a function. The Precision Bounding (PB) mode that applies
the $inexact$ function on inputs can
identify other types of instabilities, in particular catastrophic
cancellations, however, it is more difficult to apply to large code bases
as it often leads to execution errors that require human intervention.

Stochastic arithmetic quantifies numerical error using the number of significant
digits $s$ estimated among the sampled values. A common formula to determine
this number from MCA samples is presented in~\cite{parker1997monte}:
\begin{equation}
    s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a
variable sampled with MCA. This can be seen as the ratio of signal over noise.
Sohier et al.~\cite{sohier2021confidence} recently provided a generalization of
this formula to include confidence intervals.

@@

Monte Carlo Arithmetic samples the floating-point variability of a given
program, and hence the more MCA repetitions we have, the more accurate the
distribution will be. In theory, an infinite number of MCA repetitions
should give the exact distribution. While in practice low sample sizes (3 to 30)
can give good estimates, the issue is to capture ``rare events" that could make the
computation strongly diverge from the mean. Determining the number of
MCA samples to use for a given instrumented code is still an open research question.

@@

Once traces are available, \pytracer sequentially parses them and computes the
mean, standard deviation, and the number of significant digits (as in
Equation~\ref{eq:sig-digits}) of all saved values. \pytracer saves these summary
statistics in an HDF5 file, a hierarchical format that facilitates visual
browsing since only required information is loaded when requested by the user.
This operation critically relies on the ordering of call input and output
objects in the traces. It assumes that the application is
single-threaded and that the control flow does not depend on a random state.
Therefore, \mbox{\pytracer} cannot merge traces resulting from different
control flow paths and will stop the parsing when encountering such a situation.

@@

We evaluated \pytracer on two widely-used Python libraries:
SciPy~\cite{virtanen2020scipy}, the main library for scientific computing in
Python, and  scikit-learn~\cite{pedregosa2011scikit}, a reference
machine-learning framework that implements a wide range of state-of-the-art
algorithms. For each library, we executed the software examples provided in
their respective GitHub repositories. We traced each example five times, having
activated MCA through Fuzzy in the Python interpreter as well as BLAS, LAPACK, NumPy,
SciPy, and scikit-learn. These libraries are written in large parts in a
compiled language other than Python (such as C or Cython) and their execution is not 
perturbed by the sole use of Fuzzy Python. We chose to activate MCA in these libraries 
to provide an overall picture of numerical uncertainty in the tested examples. 
Depending on the use case of interest, one might choose to use MCA only in 
a subset of these libraries or even just in the Python interpreter. We
used MCA in Random Rounding (RR) mode, with the default virtual
precision in Verificarlo, namely 24 bits for single-precision floating-point
values and 53 bits for double-precision values, which simulates machine error.
We used Python-3.8.5, NumPy-1.19.1, SciPy-1.5.4, and scikit-learn-0.23.2.

\begin{table*}[]
    \centering
    \normalsize
    \begin{subfigure}[t]{\linewidth}
        \centering
        \begin{tabular}{|lll|c|c|c|}
            \hline
            \multicolumn{3}{|c}{ \multirow{2}{*}{Application} }                 & \multicolumn{3}{|c|}{Random Rounding (RR)}                                                                                                                         \\
            \cline{4-6}
                                                                                &                                                                           &                                                          & trace  & parse    & sigbits \\
            \hline
            \multicolumn{1}{|c|}{ \multirow{20}{2em}{ \rotatebox{90}{SciPy} } } & 
            \multicolumn{1}{c|}{ \multirow{3}{2em}{ \rotatebox{90}{FFT} }}      & \discreteCosineRF                                                         & \valid                                                   & \valid & \s{52}             \\ 
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \fftOneDRf                                               & \valid & \valid   & 41      \\
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \fftTwoDRf                                               & \valid & \valid   & \s{52}  \\
            \cline{2-6}
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{  \multirow{5}{*}{ \rotatebox{90}{Interpolation} } }

                                                                                & \interOneDRf                                                              & \valid                                                   & \valid & 51                 \\
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \multiRf                                                 & \valid & \valid   & \s{51}  \\
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \bsplineRf                                               & \valid & \valid   & 10      \\
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \splineOneDRf                                            & \valid & \cross   & \s{52}  \\
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{}                                                     & \splineTwoDRf                                            & \valid & \valid   & 10      \\
            \cline{2-6}
            \multicolumn{1}{|c|}{}                                              & \multicolumn{1}{c|}{ \multirow{12}{2em}{ \rotatebox{90}{Optimization} } }

@@

@@

        \pytracer trace parsing and aggregation, \emph{result}: result precision
        in number of significant bits. \valid: successful run, \cross: numerical error (tracing) or 
        different control flow paths (parsing). \textit{nan} stands for (\texttt{Not-A-Number}). 
        Example names are linked to their source code.}

@@

Table~\ref{tab:pytracer_results_summary} summarizes \pytracer's outcome on
SciPy and scikit-learn examples. Columns \textit{trace} and \textit{parse}
report on the tracing and aggregation steps with two possible outcomes:
green \valid\xspace for success and red \cross\xspace for error. Tracing errors
occur in the application as a result of 
numerical perturbations, whereas parsing errors occur when traces result from different 
control flow paths.
Overall, the tracing and parsing of SciPy and scikit-learn examples in Random
Rounding mode was entirely successful for 32/40 examples.

Column \textit{sigbits} in Table~\ref{tab:pytracer_results_summary} summarizes
the result precision for the main outputs of SciPy and scikit-learn examples. We
measured the result precision in significant bits on the final primary example output
by taking the element-wise mean for non-scalar values and, if the example
produced multiple outputs, the average value among them. Significant bits were computed on the final program output
and could therefore be computed even when parsing failed. We use the term ``result precision"
to denote the statistical variability measured in the number of significant bits, in 
contrast with the other meanings of the word precision used in the expressions
``virtual precision" or ``single/double precision'' in this paper. The numerical
quality across all examples varied from 0 to 52, with an average of 44
significant bits. We note that only one example failed due to a
numerical error (\texttt{face\_recognition}
for RR and \texttt{Bayesian Ridge Regression} for full MCA).

@@

PyTracer supports native floating-point formats as well as compound
data types, such as complex floating-point values, N-dimensional arrays,
and arbitrary Python classes. Figures~\ref{fig:fft1D_inputs}
and~\ref{fig:fft1D_outputs} show the visualization of the
\texttt{fft\_1D} input and output values. Results are complex values
encoded in double precision. PyTracer allows visualizing the
result precision of real (blue points) and imaginary (orange
points) parts. The top row in Figures~\ref{fig:fft1D_inputs}
and~\ref{fig:fft1D_outputs} show the input mean value while the bottom
row shows the standard deviation. Even though Random Rounding
(RR) was used, function inputs were perturbed since they result from
multiple intermediate computations perturbed with RR. The points with
low magnitude in Figure~\ref{fig:fft1D_inputs} correspond to inputs
near 0 when the input of $\sin$ or $\exp$ is close to $k\pi$, $k \in
    \mathbb{Z}$. Indeed, since MCA introduces noise, the result is not
exactly 0. We can see in Figure~\ref{fig:fft1D_inputs} that the maximal
noise introduced by MCA on the inputs is in the order of $10^{-14}$,
two orders of magnitude higher than the $ulp \simeq 10^-16$ for double precision. This slight
noise introduced can be interpreted as white noise on the input signal. We can
see on the bottom row of Figure~\ref{fig:fft1D_outputs} that the frequency noise
is of the same order of magnitude as the input noise, which is expected. The
two peaks at x=38 and x=562 correspond to the actual frequencies of the input
signal.

Figure~\ref{fig:spline2d_rr} shows the results of the spline evaluation with
\texttt{bisplev} from the \texttt{spline\_2D} example. It computes the 2D spline
interpolation for function $z=(x+y)e^{-6(x^2+y^2)}$, with $(x,y) \in
    [-1,1]\times[-1,1]$. The result is a 2D array that is natively supported by
PyTracer. We observe a significant loss of result precision in some regions, which follows
an interesting pattern that could be further investigated.
\add{This example shows that \pytracer workds on 2D inputs and the importance of
    visualizing complex data and not only reporting the values.}

@@

We illustrate on two examples how PyTracer is effective for analyzing
numerical stability in complex codes at a fine-grained level: Face recognition and Bayesian Ridge Regression.
The face recognition example uses eigenfaces (eigenvectors) and Support Vector
Machines (SVMs) to classify faces from the Labeled Faces in the Wild dataset.
This example uses a Principal Component Analysis (PCA) to extract the eigenfaces
and then trains an SVM classifier to predict the label associated with a face.
When executing in PyTracer, this example crashed because the Singular Value Decomposition (SVD) did not converge.
This example is the only one to raise an error due to the
non-convergence of the SVD performed in the PCA.
The PCA uses the Randomized SVD~\cite{halko2011finding} method
that first computes a low-rank approximation of the input matrix using a
stochastic algorithm and then uses this approximation as input for the SVD. 
When walking through the code with PyTracer, we observed that
the \texttt{randomized\_svd} scikit-learn function uses the Divide \& Conquer SVD
method \texttt{gesdd}. The LAPACK library has two main methods for computing the SVD:
\texttt{gesdd} and \texttt{gesvd}. The former uses a Divide \& Conquer approach,
while the latter uses a QR decomposition. While both methods are expected to
have the same accuracy~\cite{nakatsukasa2013stable}, the Divide \& Conquer
does not converge with MCA.
